The art of modelling the 3D structure of the genome\\ Insights in {\em P. falciparum} genome organization

### 2017-07-26 doc/manuscript/paper.tex
small rewording

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwujekwe:do}.
Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host. \sout{The human host is infected with
\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
start a two weeks multiplication process. They are then released as
\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
parasites then start their ``erythrocytic'' cycle (through \textit{rings},
\textit{trophozoites} and \textit{schizonts} stages), via another round of
replications in red blood cells, through an unusual process of cell division
called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
replication and division into 13 to 32 daughter cells, until the red blood
cell bursts and provokes the release of \textit{merezoites} and the infective cycle
starts anew. This asexual replication cycle is responsible for the symptoms
and the complications of the disease: anemia, tertian fever, \dots}

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}. Thus,
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

\fixme{contrast model-driven studies with data-driven studies}

@@

In addition of having subtle differences in the derivation of wish distances,
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
summarizes the characteristic of each method.

@@

The last category of data-driven methods to infer the 3D architecture of the
genome rely on a new protocol to probe single-cell for their 3D structures
\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its early days,
and only a handful of data sets are today publicly available. The contact maps
originating from these data sets are very sparse, and specific methods to
infer 3D structures need to be developed specifically for sc-HiC. The
challenge of constructing models from this data has been tackled very
differently by two groups.

@@

\begin{table}[ht!]
\begin{center}
\scriptsize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{peng:sequencing}} & autochrom3D & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\scriptsize{\cite{trieu:large}} & & C & & & \\
\scriptsize{\cite{trieu:3D}} & & C & &  & \\
\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

@@

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
are constrained into a spherical ball representing the nucleus; \textit{(ii)}
centromeres are constrained into a spherical ball tethered to the nuclear
membrane; \textit{(iii)} telomeres are tethered to the nuclear membraine;
\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
spherical ball opposite to the centromeres.

Let us take a closer look at the model proposed by \citet{tjong:physical}.
The 16 chromosomes are modeled as beads-on-a-string spaced 3.2kb apart. The
authors cast an optimization problem including the following restraints:

\begin{itemize}
\item \textbf{Adjacent Beads Constraints} Two adjacent beads spaced 3.2kb
apart are constrained to be 30~nm apart;
\item \textbf{Volume Exclusion Constraints} Each pairs of beads are spaced at
minimum $30~nm$ apart;
\item \textbf{Nucleus Constraint} All beads lie in a spherical nucleus of
radius $1\mu m$
\item \textbf{Centromeres Constraints} Centromeres are localized in a small sphere of
50nm abutting the nuclear envelop.
\item \textbf{Telomere Constraints} Telomeres are located at most 50nm from the nuclear envelop.
\item \textbf{rDNA restraint} The repeated rDNA located on chromosome 12 is
constrained to be inside the nucleolus;
\item \textbf{Nucleolus exclusion restraint} All beads that isn't rDNA is
constrained outside of the nucleolus;
\item \textbf{Chain stiffness}: an angular restraint is imposed on three
consecutive beads to reproduce chain stiffness:
\begin{equation}
\frac{1}{2} k_{\text{angle}} \sum^{N - 2}_{i = 1} \left( 1 - \frac{x_{i + 1} -
x_i}{\|x_{i + 1} - x_i\|} \cdot \frac{x_{i + 2} - x_{i + 1}}{\|x_{i + 2} -
  x_{i + 1}\|} \right)^2\,,
\end{equation}
\end{itemize}

>>>>

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwujekwe:do}.
Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host.

\begin{figure}
\caption{The life cycle of {\em P. falciparum}}{
\small
The human host is infected by \textit{sporozoites} through the bite of a infected female  {\em Anopheles}
mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
start a two weeks multiplication process. They are then released as
\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
parasites then start their ``erythrocytic'' cycle (through \textit{rings},
\textit{trophozoites} and \textit{schizonts} stages), via another round of
replications in red blood cells, through an unusual process of cell division
called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
replication and division into 13 to 32 daughter cells, until the red blood
cell bursts and provokes the release of \textit{merezoites} and the infective cycle
starts anew. This asexual replication cycle is responsible for the symptoms
and the complications of the disease: anemia, tertian fever, \dots}
\end{figure}

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

@@

In addition of having subtle differences in the derivation of wish distances
(which can have important effects on the resulting structures),
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
summarizes the characteristic of each method.

@@

The last category of data-driven methods to infer the 3D architecture of the
genome rely on a new protocol to probe single-cell for their 3D structures
\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its
early days, and, despite potential for assessing the variability of
cell-to-cell genome architecture in a genome-wide fashion, only a handful of
data sets are today publicly available. The contact maps originating from
these data sets are very sparse, and specific methods to infer 3D structures
need to be developed specifically for sc-HiC. Two methods have been developed
to tackle the challenging task of constructing models from this data.

@@

\begin{table}[ht!]
\begin{center}
\scriptsize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{peng:sequencing}} & autochrom3D & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\scriptsize{\cite{trieu:large}} & & C & & & \\
\scriptsize{\cite{trieu:3D}} & & C & &  & \\
\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{\small In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

@@

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
are constrained into a spherical ball representing the nucleus; \textit{(ii)}
centromeres are constrained into a spherical ball tethered to the nuclear
membrane; \textit{(iii)} telomeres are tethered to the nuclear membrane;
\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
spherical ball opposite to the centromeres. A scoring function eng
